Type Error Telemetry at Scale: Pseudonymized and Stochiastic

### 2023-09-17 programming23/paper.tex
example timelines
compress up/down sessions into counts
talk about timestamps, cleaning

IDEs use telemetry in a similar fashion to most user-facing
applications, for example VSCode~\cite{vsc-telemetry} and
IntelliJ~\cite{intellij-telemetry} report telemetry.

@@

\subsection{Telemetry records}

@@

In order to avoid swamping our telemtry servers, we randomly throttle
the generation of telemetry records:
\begin{itemize}
  \item
    1\% of Roblox Studio sessions were enrolled in generating telemetry, and
  \item
    0.5\% of runs of the typechecker (approximately 1 in 200 keystrokes)
      in an enrolled session generated a telemetry record.
\end{itemize}
In order to find how file-switching correlated with type errors, we additionally have:
\begin{itemize}
  \item
    every run of the typechecker in an enrolled session which switched file generated a telemetry record.
\end{itemize}
Each telemetry record contains:
\begin{description}
  \item[D1] metadata: a psedonymized session identifier, timestamp and session duration,
  \item[D2] mode of the current file: (\mnocheck{}, \mnonstrict{}, or \mstrict{}),
  \item[D3] reason for sending: (random selection, or file switch),
  \item[D4] size of codebase: number of files, and number of lines of code type checked,
  \item[D5] size of edit range
  \item[D6] overall type errors: for previous and current states, we have the total number, the nunber in the current file, and number in the edit range,
  \item[D7] specific errors in the edit range: for the previous and current states, for each type error code, the number type errors in the edit range,
  \item[D8] forced strict errors: specific errors for the
    edit range computed as if the user had strict mode enabled.
  \item[D9] TooComplexErrors: project-wide count for this one specific error type
\end{description}
To track the edit range, we record a start and end position, which we
update appropriately on every edit. This can result in very large edit
ranges (for example if the user edits an the beginning and end of the
file) but fits in a fixed-size telemetry record.

\subsection{what to study}

RQs + Analysis

\begin{itemize}
  \item How often does the typechecker give up with CodeTooComplex errors?
  \subitem [[ For context: the type system is designed to give useful feedback
    to all sorts of code, whether strict-typed, nonstrict-typed, or \mnocheck{}.
    But it has an escape hatch: CTC. Hope it's rarely needed. ]]
    \begin{itemize}
      \item Count the overall CTC errors (in each mode) for a coarse view of
        how rare they are. [D9 of course, D8 to compare \mnocheck{} code with
        strict-mode code via the same typechecker]
      \subitem Note: We can't look at the code to learn why the errors
        happened. Can only look for trends and tell the Roblox team where to
        look.
        %% \subitem Count CTC errors within each session to see if these errors are intermittent, or if they stick around. [D9 for the count, D8 for \mnocheck{} vs strict, D7 because CTC might show up in the edit range]
    \end{itemize}
  \item How helpful are highlighted ranges?
    \begin{itemize}
      \item How often do edits in a highlighted range remove type errors? How
        often do they add new ones?
      \subitem [D7 compare specific old vs specific new]
      \subitem [may want to filter using D2 + D5]
      \item How often do edits outside the range remove type errors? How often do they add new ones?
      \subitem [D6 looking for old count vs. new count difference without a matching delta in the D7 counts]
      \item Which specific errors seem easy / hard to fix via highlights?
      \subitem [D6 removed by edit => easy; D6 survive edit => hard]
    \end{itemize}
  \item In what ways do types improve quality of life?
  \begin{itemize}
    \item Does \mnocheck{} code have many more latent errors than the others?
    \subitem [D8 latent errors come from the "what if" analysis]
  \end{itemize}
\end{itemize}

% What to study?
% 
% Sanity checks
% How large are edit ranges? Plot the distribution.
% How long between logs? Long times & very short times are problematic.
% General Roblox trends
% How many sessions use strict / nonstrict / \mnocheck{}?
% Expect few strict/nonstrict, lots of \mnocheck{}
% How often does mode change within a session?
% Expect few of these b/c error reports are easy to ignore; non-intrusive
% If a down-switch happens, expect the current #errors to be high
% If an up-switch happens, hope to stay up
% How many CodeTooComplex errors arise, overall?
% Expect very few. Hopefully zero.
% How often do module switches happen?
% Expect not-too-many, with some logs in between each time
% Compare strict vs. nonstrict vs. nocheck
% Get median # errors across all snapshots.
% Expect nocheck median to be higher than the others.
% Plot #errors vs time, or vs codebase size
% Expect nocheck errors to keep going up. For other modes, #errors should stay low.
% General type error trends
% Plot the distribution of #errors across all edit ranges
% Expect different counts based on error-kind. SEE BELOW
% Count specific errors that get fixed. For each kind of error, how many times is it old-in-range but not new-in-range? Compare to #old-in-range totals.
% Expect all errors to have a decent fix rate.
% If one error has a lower % than others, its error range may be misleading. Or, maybe it's hard to fix
% Type errors vs. edit range
% Compare #old vs. #new in range
% Expect a decrease, #old > #new
% If there's an increase, look for an explanation. Could be a huge edit range. Could be a 1-line range that we captured 2x in a row. Could be that this user is simply ignoring type errors.
% Compare [#old-in-range vs. (total #old - total #new)]
% If #old-in-range = 0, then hope the total # increases
% Decrease => nonlocal edits fixed an error => type error hilite may be inaccurate
% If #old-in-range > 0, then hope the total # decreases
% Increase (in range or otherwise) => made things worse

\section{Predictions}
%%bg: merge with prev section, on telemetry design?

nocheck forcestrict keep growing

nonstrict te tend to zero, fixing your program should fix the errors, no false positives

nonstrict fs unclear

\section{Data Cleaning}

Deduplicated records with the same timestamp,
kept only the first.
Timestamps by millisecond.

Records come with two timestamps, a \emph{client timestamp}
that the IDE generates before sending the record
and a \emph{server timestamp} that the telemetry server
generates when it receives the record.
We collected timestamps to sort records from earliest to latest.
The server timestamps, however, are less useful for sorting
because the server processes groups of records in a batch and
often assigns equal timestamps to dozens of records.
Hence, we ignore the server timestamps.
Throughout the paper a \emph{timestamp} is always a client-side timestamp.

Edit ranges are negative in a handful of records (1,533).
%% out/negative-edit-range.txt
%% actual value is NOT negative, but large positive over 4 billion
%% example: 4294967293
This may happen when a user deletes a large block of code.
Since the issue affects so few records, we omit those ranges
from the analysis.

Sessions, ignore if too close to start or end.
Deciding if a session is close is tricky because we don't have
a firm start and end time.
Um, does this make sense at all?
Well anyway we took the greater number from mean vs median
gap (over all points, not just keystroke points)
and multiplied by Nx to test closeness to a boundary.
With N=2 we drop 1 nonstrict session, 0 strict (not sure about nocheck).
With N=4 we drop 3, with N=8 we drop 9.

FILL remove all syntax errors from later stages?

\section{Results}
\label{s:data}

\begin{figure}[t]
  %% TODO bigger text
  %% code/row-distribution.rkt
  \includegraphics[width=\columnwidth]{img/row-distribution.pdf}
  \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California. Shaded ranges correspond to weekends.}
  \label{f:records-per-hour}
\end{figure}

FILL investigate one tower

@@

\Cref{f:records-per-hour} shows when data arrived across the whole dataset.
FILL: currently clientTimestamp, which has caveats.
Unclear what timezone they are from.
Some records have a client timestamp much older than the server, up to 1 week behind.
Some are ahead of the server by a day.
Why?
Could be wrong local settings on the computer, or a delayed record due to lost
internet connection or closed laptop.

@@

@@

@@

\begin{figure}[t]
  \mnocheck{}

  \mnonstrict{}

  \mstrict{}

@@

TODO individual sessions

>>>>

IDEs use telemetry in a similar fashion to most user-facing
applications, for example VSCode~\cite{vsc-telemetry} and
IntelliJ~\cite{intellij-telemetry} report telemetry.
See~\cref{s:related} for futher comparisons among telemetry designs.

@@

\subsection{Telemetry records}
\label{s:telemetry-records}

@@

In order to avoid swamping our telemetry servers, we randomly throttle
the generation of telemetry records:
\begin{itemize}
  \item
    1\% of Roblox Studio sessions were enrolled in generating telemetry, and
  \item
    0.5\% of runs of the typechecker (approximately 1 in 200 keystrokes)
      in an enrolled session generated a telemetry record.
\end{itemize}
In order to find how file-switching correlated with type errors, we additionally have:
\begin{itemize}
  \item
    every run of the typechecker in an enrolled session which switched file generated a telemetry record.
\end{itemize}
Each telemetry record contains:
\begin{description}
  \item[D1] metadata: a psedonymized session identifier, timestamp and session duration,
  \item[D2] mode of the current file: (\mnocheck{}, \mnonstrict{}, or \mstrict{}),
  \item[D3] reason for sending: (random selection, or file switch),
  \item[D4] size of codebase: number of files, and number of lines of code type checked,
  \item[D5] size of edit range
  \item[D6] overall type errors: for previous and current states, we have the total number, the nunber in the current file, and number in the edit range,
  \item[D7] specific errors in the edit range: for the previous and current states, for each type error code, the number type errors in the edit range,
  \item[D8] forced strict errors: specific errors for the
    edit range computed as if the user had strict mode enabled.
  \item[D9] TooComplexErrors: project-wide count for this one specific error type
\end{description}
To track the edit range, we record a start and end position, which we
update appropriately on every edit. This can result in very large edit
ranges, for example, if the user edits at the beginning and end of the
file.
The main benefit of this strategy is that it ensures constant-size telemetry records.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%  STOP!  ROUGH TEXT BELOW     %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The Data}
\label{s:data}

Roblox Studio collected type error telemetry in Spring 2023,
between February and April.
Every Studio session had a small random chance of generating
telemetry~(\cref{s:telemetry-records}).
The chosen sessions generated a record whenever the user switched modules and
randomly on each keystroke.

\Cref{f:records-per-hour} provides a time-ordered distribution of the data.
Each vertical bar counts the number of records generated per hour according to
a client-side timestamp.
The bars are labeled with a California time zone because that is the locale of
the Roblox servers, which parsed the timestamps.
Most buckets contain roughly one thousand records, but some have very few
and a handful of others exceed three thousand.
%% \FILL{} investigate one tower, what's happening?? all one session?
There is a sharp drop in early April because we disabled telemetry at that point
but some clients continued to generate data
(\FILL{} Q. Alan, is that an ok sentence? I'm not sure why there's a long tail,
guess it could be people not updating studio or people with strange time settings
on their machine.)
We do not know which timezone the records originated in, but since the counts
tend to peak midday California time it seems that the majority
of Roblox developers are following a Western Hemisphere schedule.
%% \FILL{} probably North America, but do we really want to guess?
The counts for weekends (shaded regions) are often the tallest; these days presumably
have a larger number of sessions for telemetry to observe.

\begin{figure}[t]
  %% TODO bigger text
  %% TODO weird artifact (|) in x-min label
  %% code/row-distribution.rkt
  \includegraphics[width=\columnwidth]{img/row-distribution.pdf}
  \caption{Telemetry records per hour. Each tick on the $x$-axis marks the start of a new day in California. Shaded ranges correspond to weekends.}
  \label{f:records-per-hour}
\end{figure}

\subsection{Data Cleaning}
\label{s:data-cleaning}

A close inspection of the data revealed a few anomalies.
For the most part, we omit the strange data from futher analysis.

First, some records have odd timestamps according to the server's records.
Every telemetry record comes with two timestamps: a client-side timestamp
from when Roblox Studio created it and a server-side timestamp from when
the server posted it to its database.
The server timestamps are little use for our analysis because they bunch
together; it is common for several records to have identical server
timestamps.
However, server timestamps are useful for auditing because they should
be relatively close in value to the client timestamps.
For some records (<100? \FILL{}), though, the client is off by a large
margin: up to one week behind and up to one day ahead of the server.
We attribute the large delays to issues when sending telemetry (lost internet
connection, etc.) and smaller offsets to incorrect time settings on the client
machine.
In any event, we do not filter these records from further analysis.

Second, some records have identical (client) timestamps.
This can happen legitimately if type analysis runs several times
in the same millisecond, but it is unlikely that a human would
enter keystrokes or switch modules quickly enough.
We keep the first record with a distinct timestamp and discard the others.
%% TODO wait, better deduplicate by session and not globally!!!

%% out/negative-edit-range.txt
%% actual value is NOT negative, but large positive over 4 billion
%% example: 4294967293
Third, the edit ranges are negative in a few hundred records~(1,533).
This is likely due to the user deleting a large block of code.
Since the issue affects so few records, we omit the negative ranges
from analysis (but, we do use other parts of the records; for example
the timestamps influence~\cref{f:records-per-hour}).
(\FILL{} Alan, double-check reasoning)

\subsection{Codebase Size}
\label{s:codebase-size}

@@

TODO findings lessons etc.

@@

\subsection{??? records}

@@

wow 619 up-pairs, 583 down-pairs
=== UP
min -3 max 57 median 0 mean 2.62 stddev 6.8390518952707335
=== DOWN
min -48 max 30 median 0 mean -0.31 stddev 4.498748355271854

@@

\begin{figure}[t]\centering
  %% TODO can we 
  \mnonstrict{}\\
  \includegraphics[width=\columnwidth]{img/timeline-nonstrict.pdf}

  \medskip{}
  \mstrict{}\\
  \includegraphics[width=\columnwidth]{img/timeline-strict.pdf}

@@
